---
title: "RMS Titanic"
author: "Laurel Cox"
date: 2026-9-6
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

  - There are 5 columns: class, sex, age, survived, and n (number of people in that category)
  - There are 4 classes: 1st, 2nd, 3rd, and Crew

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
  - There are 2201 passengers in this dataset
- Are there any differences?
  - Our total is 2201 but on Wiki it's 2208
- If yes, what might account for those differences?
  - There seems to be ambiguity around the exact number of passengers with one reason being some people traveled under aliases which made it difficult to pin down the exact people who were on the ship. 
  - Additionally, according to [Encyclopedia Titanica](https://www.encyclopedia-titanica.org/), there were also musicians and shipyard crew who traveled as passengers but had NA values for their fares, which could have filtered them out while defining what constitutes a passenger. 

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  ggplot(aes(x = Sex, y = n, fill = Class)) +
  geom_col(position = "Dodge") +
  labs(
    title = "Count for Class and Sex",
    y = "n (count)"
  )
```

**Observations**:

- A lot more men were present on the ship then women, one reason being that most of the crew is made up of men and accounted for almost half of the passengers.
- First class is the only class where women outnumbered guys.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  ggplot(aes(x = Sex, y = Prop, fill = Class)) +
  geom_col(position = "Dodge") +
  labs(
    title = "Proportional Count for Class and Sex",
  )
```

**Observations**:

- Write your observations here.
  - 1st and 2nd class for both sex are 100% proportional. 
- Is there anything *fishy* going on in your plot?
  - There's a lot of fishy things going on in my plot. Namely, the plot doesn't make much sense since it shows that 1st and 2nd class for both sex are 100% proportional which isn't true. 

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  ggplot(aes(x = Sex, y = Prop, fill = Class)) +
  geom_col(position = "Dodge") + 
  facet_grid(~Age) +
  labs(
    title = "Survival by Class, Sex, and Age "
  )
```

**Observations**:

- All 1st and 2nd class children for both genders survived.
- Most of the women in 1st and 2nd class survived.
- The only children who didn't survive were in 3rd class.
- Men were more likely not to survive.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - Because Prop tells use about the survival rate between class, sex, and age groups we must use those variables to decipher what the plot tells us. My plot in q4 doesn't compare age, so the proportional values show that 1st and 2nd class for both sex are 100%. However, my new plot reveals that there were non-survivors in the 1st and 2nd class.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
