---
title: "Gapminder"
author: "Laurel Cox"
date: 2026-09-28
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- Write all variable names here
  - country, continent, year, lifeExp, pop, gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% 
  arrange(year) %>%
  pull(year) %>%
  last()

year_min <- gapminder %>% 
  arrange(year) %>%
  pull(year) %>%
  first()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  filter(gdpPercap < 90000) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot()

```

**Observations**:

- Oceania has the largest GDP per capita while Africa has the smallest.

- Oceania also has the smallest spread in GPD per capita, a possible factor could be because it has the fewest number of countries compared to the other continents.

- Europe's box plot has the widest spread.

**Difficulties & Approaches**:

- Write your challenges and your approach to solving them
  - When I first plotted my box plot with only the year filtered out, my y-axis was extremely large, making the whisker plots hard to read. To solve this, I also filtered out the largest outlier in Asia.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder_outliers <- gapminder %>%
  filter(year == year_min) %>%
  filter(gdpPercap > 12500)

gapminder_outliers
```

**Observations**:

- Identify the outlier countries from q2
  - The outlier countries with the highest GDP per capita were Kuwait, Switzerland, and United States.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter((year == year_min) | (year == year_max)) %>%
  ggplot(aes(continent, gdpPercap)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United States", "Kuwait", "Switzerland")),
    mapping = aes(color = country),
    size = 2
  )
```

**Observations**:

- Kuwait has the largest gap in GDP per capita between the max and min year.
- The US and Kuwait in their max and min years are both still above the box plot. Whereas, Switzerland's min year falls closer to Europe's median.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(country %in% c("United States", "Kuwait", "Switzerland")) %>%
  ggplot(aes(year, gdpPercap, color = country)) +
  geom_point() +
  geom_line() +
  labs(
    title = "GDP per capita over time for Kuwait, Switerland, and US"
  )
```

- How do the GDP gaps between the 3 outliers identified in q2 compare?
  - Kuwait shows a huge drop in GDP per capita between the years 1970-1980. Out of the 3 outliers, it's the only country to have dropped dramatically. Switzerland and US both shows a steady rise in GDP per capita.
  - Despite having the lowest GDP per capita at approximatly 1970, Kuwait has the highest GDP per capita of the 3 countries for the most recent year and is more volatile than Switzerland and US.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  group_by(continent, country) %>%
  summarise(avg_lifeExp = mean(lifeExp)) %>%
  ggplot(aes(continent, avg_lifeExp)) +
  geom_boxplot() +
  labs(
    title = "Average Life Expectancy of Countries Per Continent"
  )

gapminder %>%
  group_by(continent, country) %>%
  summarise(avg_lifeEXp = mean(lifeExp)) %>%
  filter(continent == "Oceania")
```

- Is Europe the prosperous powerhouse I learned about in middle school history class because of their "guns, germs, and steel"?
  - Interestingly, we can see the continent with the highest and most consistent average life expectancy is actually Oceania. When looking at the countries in Oceania, Australia and New Zealand both have average life expediencies around 74 years. Europe only falls slight behind Oceania in life expectancy by a year or less, but as I expected, the next closest continent has more than a 5 year gap.
  - [This article](https://www.psu.edu/news/research/story/australia-offers-lessons-increasing-american-life-expectancy) cites many factors to why Australia has better life expectancy compared to similar countries in the last 30 years, including gun reforms, good public transit, and reducing preventable causes of death across age ranges.

```{r q5-task3}
## TASK: Your third graph
asia_outlier <- gapminder %>%
  group_by(country) %>%
  mutate(avg_lifeExp = mean(lifeExp)) %>%
  filter(continent == "Asia", avg_lifeExp < 40)

asia_outlier %>%
  ggplot(aes(year, lifeExp)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Afganistan's Life Expectancy over Time"
  )
```

- Which country had the lowest average life expectancy?

  - The lowest point from my 2nd graph in q5 was in Asia, so after some manipulation Afghanistan was revealed to be that point.

  - In the early 1950's, Afghanistan had a life expectancy of below 30 years. Up until the 1990s, there was steady improvement to life expectancy - getting to about 41 years. But then it plataues until the 2000s. This is most likely explained by the reign of the Tailban and the resulting destabilization of healthcare.
